SDSS-IV MaNGA: The Universal Profile of Enhanced Star Formation in Close Galaxy Pairs

### 2020-08-19 maintex.tex
Next draft for submission to MaNGA team

\title{
SDSS-IV MaNGA: Mass-Independent Star Formation Enhancement in Close Galaxy Pairs
}

@@

\begin{abstract}
We study the spatial profile of the star formation rate (SFR) in a sample of 220 interacting galaxies within the SDSS-IV MaNGA survey. We find that the specific star formation rate (sSFR) is centrally (e.g. within a single effective radius) enhanced in interacting galaxies by $\sim$0.25 $\pm$ 0.1 but that there is no enhancement or suppression in their disks. We split the radial profile into five separate stellar mass bins from \logm\ $=$ 9.0$-$11.5 and find that the sSFR enhancement has no dependence on the stellar mass of the paired galaxies. Within one effective radius, the sSFR enhancement in each of the mass bins falls within $\sim$0.1 dex of the median profile for the whole pair sample. We then use a second method to compare paired galaxies and control galaxies where each paired galaxies is matched to a set of 20 control galaxies based on stellar mass and redshift. The sSFR enhancements found are still mass independent, reinforcing the idea that low mass galaxies experience the same level of enhancement to their star formation rate as high mass galaxies. Since the merger induced sSFR enhancement is constant over stellar mass, low mass galaxies will experience a higher fractional mass growth rate of 1$-$6\%, assuming a constant burst merger induced star formation over a 1 Gyr period. This indicates that merger induced star formation has a significant impact on the mass evolution of low mass galaxies.

@@

This star formation enhancement increases with closer projected separations between the two galaxies \citep{Ellison:2008, Scudder:2012} but has still shown to be detectable out to projected separations of 150 kpc \citep{Patton:2013}. The level of the star formation enhancement also increases with smaller stellar mass ratios, $|\Delta$log($M$)$|$, between the two galaxies \citep{Ellison:2008}. Similar mass galaxies show the strongest sSFR effects while large central galaxies interacting with small satellite galaxies show weaker sSFR effects. Because of this, most surveys will split their sample into major mergers, $|\Delta$log($M$)$|$ $\le$ 0.5, and minor mergers, 0.5 $\le$ $|\Delta$log($M$)$|$ $\le$ 1.0, where major mergers are expected to show the strongest interaction induced effects.

@@

Where NUV$-$r is the color and M$_i$ is the i-band magnitude from the NSA catalog. We have checked the selected star forming galaxies on the BPT diagram \citep{Baldwin:1981} and confirmed that there is not a significant number of AGN in our sample.

On top of the color selection cuts, we require that all galaxies are in the Primary or Secondary MaNGA subsamples \citep{Wake:2017}. The stellar mass range of the galaxy sample is between \logm\ $=$ 9.0$-$11.5.

@@

From the NSA catalog's 641,409 galaxies, we find 492 galaxies which are paired to MaNGA targets. MaNGA targets which have both an inside-IFU and an outside-IFU pair are left to the inside-IFU sample. After restricting the sample to SFGs, we have 133 MaNGA targets with paired galaxies outside of the IFU.

@@

The SFR in galaxies has been shown to be dependent on both stellar mass and redshift of the galaxies \citep{Noeske:2007}. This effect needs to be controlled for when comparing the paired and control galaxies to isolate the changes in the sSFR due to merger induced effects. For each galaxy pair we build a subset of 20 control galaxies which are matched in stellar mass and redshift.

We build a sample of isolated control galaxies from the remaining MaNGA galaxies by selecting the MaNGA target galaxies which have no spectroscopic companions within r$_p$ $<$ 50 kpc and $\Delta$v $<$ 500 km s$^{-1}$ either inside or outside of the IFU. This gives us a set of 1891 SFGs and 1528 SFGs which are covered by both \citet{Simard:2011} and \citet{Mendel:2014}.

@@

This will allow us to compare the SFR in high mass regions, like the centers of the galaxies, to the SFR lower mass regions, like the disk of the galaxies. We have also found a strong correlation between sSFR and \ewha and have found the same results using \ewha. This is useful to check as \ewha\ is an observable and independent on {\sc SPFIT}'s measurement of the stellar mass.

@@

The NSA catalog has complete coverage over the MaNGA sample (since MaNGA selects its targets from this catalog); however, it tends to fail to properly fit paired galaxies with close on-sky separations. We found that the apertures from \citet{Simard:2011} work better for these close galaxies, but the catalog does not completely cover the MaNGA sample. We use NSA catalog for the outside-IFU pair sample because they are well separated on the sky and the \citet{Simard:2011} catalog on the inside-IFU sample. In the cases where an Inside-IFU paired galaxy is not covered by \citet{Simard:2011}, we use the ellipse from the NSA catalog.

@@

We use our catalog of photometric objects and the \citet{Simard:2011} catalog to mask foreground/background objects which may contaminate the data. Objects in our photometric catalog are masked out with a 2\arcsec\ circular aperture and the galaxies from \citet{Simard:2011} are masked out with a 2.0 R/\reff\ elliptical aperture, using the r-band S\'ersic apertures from the catalog.

@@

In the second method, we match each paired galaxy to a set of 20 control galaxies of a similar stellar mass and redshift, as mentioned in Section \ref{sec:control}. We take the individual averaged profiles of the selected control galaxies and take the median value of the profiles at each radius bin giving us a stacked profile of control galaxies. The error of the profile is the standard error of the mean of the individual profiles. We also take into account the error associated with the random selection of controls using a bootstrapping method to quantify the variation in the values of the stacked profiles over 1000 iterations. This random selection contributes an error of about 0.1 dex.

@@

The difference profiles at this stage have now been controlled for stellar mass, redshift, and galaxy size. We can freely stack these profiles to study the global properties of the merging galaxies. We take all of the $\Delta$log(sSFR) profiles and take the median value within each radius bin to create a stacked profile of the individual difference profiles. We split the individual difference profiles into five evenly spaced stellar mass bins between \logm\ $=$ 9.0$-$11.5 and stack the difference profiles within each mass bin. This gives us four difference profiles covering four different stellar mass ranges shown in the left hand panel of Figure \ref{fig:ssfr_mass}.

The difference profiles are shown to be centrally enhanced by $\sim$0.25 $\pm$ 0.1 dex. The enhancement falls to zero around 1.0 \reff\ and is slightly suppressed in their disks by $\sim$0.1 dex. The profiles of the pair stellar masses of \logm\ $=$ 9.0$-$9.5 show a strong enhancement in their disks of $\sim$0.1$-$0.2 $\pm$ 0.1 dex.

@@

@@

For this estimation we use a timescale of 1 Gyr, which is the order of the typical timescale between the passage of the first and second pericenters in dark-matter simulations \citep{Boylan-Kolchin:2008}. The actual timescale for star formation enhancement is likely shorter; however, we use the 1 Gyr value as an order of magnitude guess.

@@

The sSFR enhancement increasing with closer projected separations is to be expected and has been shown in previous studies \citep{Ellison:2008, Scudder:2012, Patton:2013}. The passage of the first pericenter is predicted to trigger a burst of star formation which falls as the two galaxies separate \citep{Scudder:2012}. The sSFR enhancement in our sample persists out to 50 kpc, \citet{Patton:2013} showed that the burst of star formation can persist out to 150 kpc.

For galaxies with separations of r$_p$ $<$ 5 kpc and r$_p$ $=$ 30$-$35 kpc we find a $\Delta$log(sSFR) of zero. For the r$_p$ $<$ 5 kpc galaxy, about half of the pairs show signs of tidal features while the others do not. The pairs without the tidal features may be approaching their first pericenter, meaning that the merger-induced gas-inflows have not been triggered yet. The dips in $\Delta$log(sSFR) may also just be due to statistical error from the relatively limited sample size.

@@

@@

\begin{enumerate}
\item The sSFR is centrally enhanced within the inner 1.0 R/\reff\ by 0.25 $\pm$ 0.1 dex. 
\item The disks of the paired galaxies, R/\reff\ $=$ 1.0$-$2.5, show limited suppression in their disks, $\sim$0.1 dex.
\item The sSFR enhancement in paired galaxies is independent of the total stellar mass of the paired galaxies, Massive galaxies experience the same level of sSFR enhancement as low mass galaxies.
\item The sSFR enhancement in paired galaxies does depend on the ratio between the masses of the paired galaxies. Galaxies with small mass ratios, $|\Delta$log(M)$|$ $=$ 0.0$-$0.5, see an enhancement to sSFR of 0.3 $\pm$ 0.1 dex, which is 0.1 dex higher than the average paired galaxies in the survey. We have also showed that paired galaxies with large mass ratios, $|\Delta$log(M)$|$ $=$ 1.0, still show substantial levels of enhancement to the sSFR, 0.15$-$0.2 dex. 
\item Further, in paired galaxies with large mass ratios, the less massive galaxies experience higher levels of sSFR enhancement than the more massive galaxy in the pair, $\sim$0.1 dex.
\item The sSFR enhancement increases with closer projected separations where galaxies within 5$-$15 kpc have their sSFR enhanced by 0.3$-$0.4 $\pm$ 0.1 dex. 
\item The paired galaxies will experience a fractional mass growth rate of 1$-$6\%, where the growth in mass in largest in low mass galaxies.

>>>>

\title{
SDSS-IV MaNGA: The Universal Profile of Enhanced Star Formation in Close Galaxy Pairs
}

@@

\begin{abstract}
We study the spatial profile of the star formation rate (SFR) in a sample of 220 interacting galaxies within the SDSS-IV MaNGA survey. We find that the specific star formation rate (sSFR) is centrally (e.g. within a single effective radius) enhanced in interacting galaxies by $\sim$0.25 $\pm$ 0.1 but that there is no enhancement or suppression in their disks. We split the radial profile into five separate stellar mass bins from \logm\ $=$ 9.0$-$11.5 and find that the sSFR enhancement has no dependence on the stellar mass of the paired galaxies. Within one effective radius, the sSFR enhancement in each of the mass bins falls within $\sim$0.1 dex of the median profile for the whole pair sample. We then use a second method to compare paired galaxies and control galaxies where each paired galaxies is matched to a set of 20 control galaxies based on stellar mass and redshift. The sSFR enhancements found are still mass independent, reinforcing the idea that low mass galaxies experience the same level of enhancement to their star formation rate as high mass galaxies. We describe the profiles of the $\Delta$log(sSFR) enhancement as a universal profile because the general shape of the profile is unchanged by parameters like stellar mass, mass ratio, or projected separation. The shape of the profile is a central peak in $\Delta$log(sSFR) which linearly falls to zero around 1.0 R/\reff. Parameters like the mass ratio or projected separation may scale the height of the profile's central peak, but the general shape of the profile is preserved. Since the merger induced sSFR enhancement is constant over stellar mass, low mass galaxies will experience a higher fractional mass growth rate of 1$-$6\%, assuming a constant burst merger induced star formation over a 1 Gyr period. This indicates that merger induced star formation has a significant impact on the mass evolution of low mass galaxies.

@@

This star formation enhancement increases with closer projected separations between the two galaxies \citep{Li:2008, Ellison:2008, Scudder:2012} but has still shown to be detectable out to projected separations of 150 kpc \citep{Patton:2013}. The level of the star formation enhancement also increases with smaller stellar mass ratios, $|\Delta$log($M$)$|$, between the two galaxies \citep{Ellison:2008}. Similar mass galaxies show the strongest sSFR effects while large central galaxies interacting with small satellite galaxies show weaker sSFR effects. Because of this, most surveys will split their sample into major mergers, $|\Delta$log($M$)$|$ $\le$ 0.5, and minor mergers, 0.5 $\le$ $|\Delta$log($M$)$|$ $\le$ 1.0, where major mergers are expected to show the strongest interaction induced effects.

@@

Where NUV$-$r is the color and $\mathcal{M}_i$ is the i-band magnitude from the NSA catalog. We have checked the selected star forming galaxies on the BPT diagram \citep{Baldwin:1981} and confirmed that there is not a significant number of AGN in our sample.

On top of the color selection cuts, we require that all galaxies are in the Primary or Secondary MaNGA subsamples \citep{Wake:2017}, and that the stellar mass range of the galaxy sample is between \logm\ $=$ 9.0$-$11.5.

@@

From the NSA catalog's 641,409 galaxies, we find 492 galaxies which are paired to MaNGA targets. MaNGA targets which have both an inside-IFU and an outside-IFU pair are left to the inside-IFU sample. After restricting the sample to SFGs, we have another 133 MaNGA targets with paired galaxies outside of the IFU. This give use a total of 220 MaNGA targets in our pair sample.

@@

The SFR in galaxies has been shown to be dependent on both stellar mass and redshift of the galaxies \citep{Noeske:2007}. This effect needs to be controlled for when comparing the paired and control galaxies to isolate the changes in the sSFR due to merger induced effects.

We build a sample of isolated control galaxies from the remaining MaNGA galaxies by selecting the MaNGA target galaxies which have no spectroscopic companions within r$_p$ $<$ 50 kpc and $\Delta$v $<$ 500 km s$^{-1}$ either inside or outside of the IFU. This gives us a set of 1891 star forming control galaxies to work with.

@@

This will allow us to compare the SFR in high mass regions, like the centers of the galaxies, to the SFR lower mass regions, like the disk of the galaxies. We have also found a strong correlation between sSFR and \ewha\ and have found the same results using \ewha. This is useful to check as \ewha\ is an observable and independent of {\sc SPFIT}'s measurement of the stellar mass.

@@

The NSA catalog has complete coverage over the MaNGA sample (since MaNGA selects its targets from this catalog); however, it tends to fail to properly fit paired galaxies with close on-sky separations. We found that the apertures from \citet{Simard:2011} work better for these close galaxies, but the catalog does not completely cover the MaNGA sample. We use NSA catalog for the outside-IFU pair sample because they are well separated on the sky and the \citet{Simard:2011} catalog for the inside-IFU sample. In the cases where an Inside-IFU paired galaxy is not covered by \citet{Simard:2011}, we use the ellipse from the NSA catalog.

@@

We use our catalog of photometric objects and the \citet{Simard:2011} catalog to mask foreground/background objects within the MaNGA IFUs which may contaminate the data. Objects in our photometric catalog are masked out with a 2\arcsec\ circular aperture and the galaxies from \citet{Simard:2011} are masked out with a 2.0 R/\reff\ elliptical aperture, using the r-band S\'ersic apertures from the catalog.

@@

In the second method, we match each paired galaxy to a set of 20 control galaxies of a similar stellar mass and redshift, as mentioned in Section \ref{sec:control}. We take the individual averaged profiles of the selected control galaxies and take the median value of the profiles at each radius bin giving us a stacked profile of control galaxies. The error of the profile is the standard error of the mean of the individual profiles. We also take into account the error associated with the random selection of controls using a bootstrapping method to quantify the variation in the values of the stacked profiles over 1000 iterations. We estimate that this random selection contributes an error of about 0.1 dex.

@@

The difference profiles at this stage have now been controlled for stellar mass, redshift, and galaxy size. We can freely stack these profiles to study the global properties of the merging galaxies. We take all of the $\Delta$log(sSFR) profiles and take the median value within each radius bin to create a stacked profile of the individual difference profiles. We split the individual difference profiles into five evenly spaced stellar mass bins between \logm\ $=$ 9.0$-$11.5 and stack the difference profiles within each mass bin. This gives us five difference profiles covering five different stellar mass ranges shown in the left hand panel of Figure \ref{fig:ssfr_mass}.

The difference profiles are shown to be centrally enhanced by $\sim$0.25 $\pm$ 0.1 dex. The enhancement falls to zero around 1.0 \reff\ and is slightly suppressed in their disks by $\sim$0.1 dex. While all of the other paired galaxies show a sSFR suppression in their disks, the profiles of the pairs in the stellar mass range of \logm\ $=$ 9.0$-$9.5 show an enhancement in their disks of $\sim$0.1$-$0.2 $\pm$ 0.1 dex.

@@

Between the two different pair-control comparison methods, we conclude that the $\Delta$log(sSFR) is independent of the paired galaxies' stellar mass. Low mass galaxies are having their sSFR enhanced at a similar rate as high mass galaxies.

@@

For this estimation we use a timescale of 1 Gyr, which is the order of the typical timescale between the passage of the first and second pericenters in dark-matter simulations \citep{Boylan-Kolchin:2008}. The actual timescale for star formation enhancement is likely shorter; however, we will use the 1 Gyr value as an order of magnitude estimate.

@@

The sSFR enhancement increasing with closer projected separations is to be expected and has been shown in previous studies \citep{Li:2008, Ellison:2008, Scudder:2012, Patton:2013}. The passage of the first pericenter is predicted to trigger a burst of star formation which falls as the two galaxies separate \citep{Scudder:2012}. The sSFR enhancement in our sample persists out to 50 kpc, \citet{Patton:2013} showed that the burst of star formation can persist out to 150 kpc.

For galaxies with separations of r$_p$ $<$ 5 kpc and r$_p$ $=$ 30$-$35 kpc we find a $\Delta$log(sSFR) of zero. For the r$_p$ $<$ 5 kpc galaxies, about half of the pairs show signs of tidal features while the others do not. The pairs without the tidal features may be approaching their first pericenter, meaning that the merger-induced gas-inflows have not been triggered yet. The dips in $\Delta$log(sSFR) may also just be due to statistical error from the relatively limited sample size.

Between the three parameters, stellar mass, mass ratio, and projected separation, the projected separation has the strongest effect on the $\Delta$log(sSFR) enhancement, followed by the mass ratio. From this we see that the paired galaxies with the strongest enhancements to the central sSFR are paired galaxies with 1:1 mass ratios and  which are within 10 kpc.

Between the splitting the $\Delta$log(sSFR) profile by stellar mass, mass ratio, and projected separation, we see that the general shape of the profile within the inner 1.0 R/\reff\ is unchanged. The profile is peaked in the center of the paired galaxies and linearly (in log space) to zero around 1.0 R/\reff. Different projected separations and mass ratios may elevate or suppress the level of the enhancement in the profile's center; however, the shape of the profile is largely unchanged.

@@

\citet{Li:2008} studied the star formation enhancement over different stellar masses and saw that low mass galaxies experience greater levels of enhancement than high mass galaxies.  Galaxies in the mass range \logm\ $=$ 9.72 were shown to experience an enhancement $\sim$0.4 dex higher than galaxies in the mass range \logm\ $=$ 10.6. In this work, we find that $\Delta$log(sSFR) has no dependence on the stellar mass of the galaxy. The galaxies with stellar masses above \logm\ $\ge$ 10.5 do show slightly lower levels of enhancement compared to the lower mass galaxies; however, the pairs still do not significantly deviate from the average $\Delta$log(sSFR) of the sample.

@@

\begin{enumerate}
\item The sSFR is centrally enhanced within the inner 1.0 R/\reff\ by 0.25 $\pm$ 0.1 dex. 
\item The disks of the paired galaxies, R/\reff\ $=$ 1.0$-$2.5, show limited suppression in their disks, $\sim$0.1 dex.
\item The sSFR enhancement in paired galaxies is independent of the total stellar mass of the paired galaxies, Massive galaxies experience the same level of sSFR enhancement as low mass galaxies.
\item The sSFR enhancement in paired galaxies does depend on the ratio between the masses of the paired galaxies. Galaxies with small mass ratios, $|\Delta$log(M)$|$ $=$ 0.0$-$0.5, see an enhancement to sSFR of 0.3 $\pm$ 0.1 dex, which is 0.1 dex higher than the average paired galaxies in the survey. We have also showed that paired galaxies with large mass ratios, $|\Delta$log(M)$|$ $=$ 1.0, still show substantial levels of enhancement to the sSFR, 0.15$-$0.2 dex. 
\item Further, in paired galaxies with large mass ratios, the less massive galaxies experience higher levels of sSFR enhancement than the more massive galaxy in the pair, $\sim$0.1 dex.
\item The sSFR enhancement increases with closer projected separations where galaxies within 5$-$15 kpc have their sSFR enhanced by 0.3$-$0.4 $\pm$ 0.1 dex. 
\item When splitting the sSFR enhancement profiles by stellar mass, mass ratio, or projected separation, the general shape of the enhancement profile is unchanged. While lower projected separations or mass ratios closer to 1:1 may raise the level of the central enhancement to the profiles, the profiles still linearly fall to zero enhancement around 1.0 R/\reff\ with roughly the same gradient.
\item The paired galaxies will experience a fractional mass growth rate of 1$-$6\%, where the growth in mass in largest in low mass galaxies.
